---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```


## Loading and preprocessing the data

1. Loading the data with the following code:
```{r load and pre-process}
setwd('C:/Users/pz11965/Documents/R Coursera/Reproducible Research/Week 2/Course Project 1/RepData_PeerAssessment1-master')
unzip('activity.zip')
infile <- read.csv('activity.csv')
```

2. Convert date variable to Date class for downstream use. 
```{r}
infile$date = as.Date(infile$date)
```


## What is mean total number of steps taken per day?

Remove missing value records and perform initial diagnostics. As expected, more steps happen during the middle of the day.
```{r}
infile_no_na <- infile[is.na(infile$steps) == FALSE,]
plot(infile_no_na)
```


1. Roll up steps to day level to get the total number of steps taken per day with the following code:

```{r message = FALSE}
infile_day_level <- infile_no_na %>% group_by(date) %>% summarize(steps = sum(steps))
```

2. Create histogram of the frequency of days by step buckets

```{r}
hist(infile_day_level$steps, main="Distribution of daily steps",
        xlab = "# of steps",
        ylab = "# of days",
        ylim = c(1,30),
        breaks = 10,
        freq=TRUE
        )

```

3. Calculate the mean and median of the steps

```{r}
mean_step <- as.integer(summary(infile_day_level$steps)[4])
median_step <- as.integer(summary(infile_day_level$steps)[3])
```

The mean is `r mean_step` and the median is `r median_step`.


## What is the average daily activity pattern?

1. Time series plot of average number of steps by time interval
```{r message = FALSE}
infile_int_time_series <- infile_no_na %>% group_by(interval) %>% summarize(steps = mean(steps))
plot(infile_int_time_series, type = 'l')
max_int <- infile_int_time_series[infile_int_time_series$steps == max(infile_int_time_series$steps),]$interval
```

2. The `r max_int` interval contains the maximum average number of steps.

## Imputing missing values

```{r}
missing_records <- length(is.na(infile))
```
1. There are `r missing_records` rows that have NA steps in the data.

2. Interval is more indicative of step count than date, so we will impute missing values by using the interval mean
```{r message = FALSE}
infile_na <- infile[is.na(infile$steps) == TRUE,]
uniq_na_combo <- length(unique(infile_na$interval))
infile_uniq_no_na <- infile_no_na %>% group_by(interval) %>% summarize(steps_imputed = mean(steps))
infile_na <- merge(infile_na, infile_uniq_no_na, by = "interval")
```

3. Fill in original dataset with imputed steps for NAs

```{r}

infile_imputed <- merge(infile, infile_uniq_no_na, by = "interval")
infile_imputed[is.na(infile_imputed$steps)==TRUE,]$steps <- infile_imputed[is.na(infile_imputed$steps)==TRUE,]$steps_imputed
infile_imputed <- infile_imputed[1:3]

```

4. Create histogram and calculate mean and median
```{r message = FALSE}

infile_day_level <- infile_imputed %>% group_by(date) %>% summarize(steps = sum(steps))

hist(infile_day_level$steps, main="Distribution of daily steps from imputed dataset",
        xlab = "# of steps",
        ylab = "# of days",
        ylim = c(1,30),
        breaks = 10,
        freq=TRUE
        )

mean_step_imputed <- as.integer(summary(infile_day_level$steps)[4])
median_step_imputed <- as.integer(summary(infile_day_level$steps)[3])

```

The mean from the imputed dataset is `r mean_step_imputed`; the median from the imputed dataset is `r median_step_imputed`. The missing values do not have significant impact on the results as they do not differ much from the orignal dataset with NAs.



## Are there differences in activity patterns between weekdays and weekends?

1. Add variable for weekday/weekend qualifier

```{r}

wkdy <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
infile_imputed$day <- factor((weekdays(infile_imputed$date) %in% wkdy), 
         levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))

```

2. Average step numbers by interval for Weekend vs weekday

```{r message=FALSE}

infile_int_time_series_imputed <- infile_imputed %>% group_by(interval,day) %>% summarize(steps = mean(steps))

library(lattice)
xyplot(steps ~ interval | day,
       data = infile_int_time_series_imputed,
       type = "l",
       layout = c(1,2),
       ylab = "Number of steps",
       main = "Interval average step numbers (weekday vs weekend)")

```

There tends to be more steps distributed more evenly across the day on the weekends than weekdays. The activity during weekdays tend to happen earlier in the day.